---
title: "Informe de R en Markdown"
runningheader: "R Markdown" # only for pdf output
subtitle: "Una implementación con R Markdown" # only for html output
author: "Felipe Ruiz"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introducción

El estilo de Tufte handout es un estilo que Edward Tufte utiliza en sus libros y folletos. El estilo de Tufte se caracteriza por el uso extensivo de notas al margen, la integración ajustada de gráficos con texto y una tipografía bien establecida. Este estilo ha sido implementado en LaTeX y HTML/CSS[^1], respectivamente. Hemos portado ambas implementaciones al [**tufte** package](https://github.com/rstudio/tufte). Si deseas una salida en LaTeX/PDF, puedes utilizar el formato `tufte_handout` para folletos y `tufte_book` para libros. Para la salida en HTML, utiliza `tufte_html`. Estos formatos se pueden especificar en el encabezado YAML al comienzo de un documento R Markdown (ver el ejemplo a continuación) o pasarlos a la función `rmarkdown::render()`. Consulta @R-rmarkdown para obtener más información sobre **rmarkdown**.


[^1]: Mire el repositorio [tufte-latex](https://github.com/tufte-latex/tufte-latex) and [tufte-css](https://github.com/edwardtufte/tufte-css)

``` yaml
---
title: "An Example Using the Tufte Style"
author: "John Smith"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---
```

Este paquete tiene dos objetivos:

1. Producir tanto salida en PDF como en HTML con estilos similares desde el mismo documento R Markdown.
2. Proporcionar una sintaxis simple para escribir elementos del estilo Tufte, como notas al margen y figuras en el margen. Por ejemplo, si deseas una figura en el margen, todo lo que necesitas hacer es usar la opción del chunk `fig.margin = TRUE`, y nosotros nos encargaremos de los detalles para que no tengas que preocuparte por `\begin{marginfigure} \end{marginfigure}` o `<span class="marginfigure"> </span>`; el código LaTeX y HTML subyacente puede ser complicado, pero nunca necesitas aprender o escribir dicho código.

Si tienes alguna solicitud de función o encuentras errores en **tufte**, no dudes en informarlos en <https://github.com/rstudio/tufte/issues>. Para preguntas generales, puedes preguntar en StackOverflow: <https://stackoverflow.com/tags/rmarkdown>.


# Encabezados

Este estilo proporciona encabezados de primer y segundo nivel (es decir, `#` y `##`), como se demuestra en la siguiente sección. Puedes obtener una salida inesperada si intentas usar `###` y encabezados más pequeños.

`r newthought('En sus libros más recientes')`[^2], Tufte comienza cada sección con un poco de espacio vertical, un párrafo sin sangría y establece las primeras palabras de la oración en versalitas. Para lograr esto con este estilo, llama a la función `newthought()` en **tufte** en una *expresión R en línea* `` `r ` `` como se muestra al principio de este párrafo.[^3]

[^2]: [Beautiful Evidence](https://www.edwardtufte.com/tufte/books_be)

[^3]: Ten en cuenta que no debes asumir que **tufte** ha sido instalado a tu sesión de R. Debes llamar `library(tufte)` en tu documento R Markdown antes de llamar a `newthought()`, o utilizar `tufte::newthought()`.


# Figuras

## Figuras al margen

Las imágenes y gráficos juegan un papel integral en el trabajo de Tufte. Para colocar figuras en el margen, puedes usar la opción de chunk **knitr** `fig.margin = TRUE`. Por ejemplo:

```{r fig-margin, fig.margin = TRUE, fig.cap = "MPG vs horsepower, por tipo de transmisión.", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE}
library(ggplot2)
mtcars2 <- mtcars
mtcars2$am <- factor(
  mtcars$am, labels = c('automatic', 'manual')
)
ggplot(mtcars2, aes(hp, mpg, color = am)) +
  geom_point() + geom_smooth() +
  theme(legend.position = 'bottom')
```

Observa el uso de la opción `fig.cap` para proporcionar un pie de figura. Puedes ajustar las proporciones de las figuras utilizando las opciones del chunk `fig.width` y `fig.height`. Estas se especifican en pulgadas y se escalarán automáticamente para ajustarse al margen del folleto.


## Contenido arbitrio al margen

De hecho, puedes incluir cualquier cosa en el margen utilizando el motor **knitr** llamado `marginfigure`. A diferencia de los bloques de código R ```` ```{r} ````, puedes escribir un chunk empezando con ```` ```{marginfigure} ```` y luego colocas el contenido en el bloque. Mira un ejemplo sobre el primer teorema fundamental del cálculo a la derecha.

```{marginfigure}
Sabemos del _primer teorema fundamental del cálculo_ que para $x$ en $[a, b]$:
$$\frac{d}{dx}\left( \int_{a}^{x} f(u)\,du\right)=f(x).$$
```

Por razones de portabilidad entre LaTeX y HTML, debes mantener el contenido del margen lo más simple posible (sintaxis-wise) en los bloques `marginfigure`. Puedes usar una sintaxis de Markdown simple como texto en `**negrita**` o `_cursiva_`, pero evita el uso de notas al pie, citas o elementos de nivel de bloque (por ejemplo, citas textuales y listas) allí.

Nota: si estableces `echo = FALSE` en tus opciones globales de chunk, deberás agregar `echo = TRUE` al chunk para mostrar una figura en el margen, por ejemplo ```` ```{marginfigure, echo = TRUE} ````.


## Figuras de ancho completo

Puedes hacer que las figuras abarquen toda la página utilizando la opción del chunk  `fig.fullwidth = TRUE`.

```{r fig-fullwidth, fig.width = 10, fig.height = 2, fig.fullwidth = TRUE, fig.cap = "A full width figure.", warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds, aes(carat, price)) + geom_smooth() +
  facet_grid(~ cut)
```

Other chunk options related to figures can still be used, such as `fig.width`, `fig.cap`, `out.width`, and so on. For full width figures, usually `fig.width` is large and `fig.height` is small. In the above example, the plot size is $10 \times 2$.

## Arbitrary Full Width Content

Any content can span to the full width of the page. This feature requires Pandoc 2.0 or above. All you need is to put your content in a fenced `Div` with the class `fullwidth`, e.g.,

``` md
::: {.fullwidth}
Any _full width_ content here.
:::
```

Below is an example:

::: fullwidth
*R is free software and comes with ABSOLUTELY NO WARRANTY.* You are welcome to redistribute it under the terms of the GNU General Public License versions 2 or 3. For more information about these matters see <https://www.gnu.org/licenses/>.
:::

## Main Column Figures

Besides margin and full width figures, you can of course also include figures constrained to the main column. This is the default type of figures in the LaTeX/HTML output.

```{r fig-main, fig.cap = "A figure in the main column.", cache=TRUE}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
```

# Sidenotes

One of the most prominent and distinctive features of this style is the extensive use of sidenotes. There is a wide margin to provide ample room for sidenotes and small figures. Any use of a footnote will automatically be converted to a sidenote. [^4]

[^4]: This is a sidenote that was entered using a footnote.

If you'd like to place ancillary information in the margin without the sidenote mark (the superscript number), you can use the `margin_note()` function from **tufte** in an inline R expression. `r margin_note("This is a margin note.  Notice that there is no number preceding the note.")` This function does not process the text with Pandoc, so Markdown syntax will not work here. If you need to write anything in Markdown syntax, please use the `marginfigure` block described previously.

# References

References can be displayed as margin notes for HTML output. For example, we can cite R here [@R-base]. To enable this feature, you must set `link-citations: yes` in the YAML metadata, and the version of Pandoc should at least 2.11 or `pandoc-citeproc` should be available and at least 0.7.2. You can always install your own version of Pandoc from <https://pandoc.org/installing.html> if the version is not sufficient. To check the version of `pandoc-citeproc` in your system, you may run this in R:

```{r eval=FALSE}
system2('pandoc-citeproc', '--version')
```

If your version of `pandoc-citeproc` is too low, or you did not set `link-citations: yes` in YAML, references in the HTML output will be placed at the end of the output document.

# Tables

You can use the `kable()` function from the **knitr** package to format tables that integrate well with the rest of the Tufte handout style. The table captions are placed in the margin like figures in the HTML output.

```{r}
knitr::kable(
  mtcars[1:6, 1:6], caption = 'A subset of mtcars.'
)
```

# Block Quotes

We know from the Markdown syntax that paragraphs that start with `>` are converted to block quotes. If you want to add a right-aligned footer for the quote, you may use the function `quote_footer()` from **tufte** in an inline R expression. Here is an example:

> "If it weren't for my lawyer, I'd still be in prison. It went a lot faster with two people digging."
>
> `r tufte::quote_footer('--- Joe Martin')` Without using `quote_footer()`, it looks like this (the second line is just a normal paragraph):

> "Great people talk about ideas, average people talk about things, and small people talk about wine."
>
> --- Fran Lebowitz \# Responsiveness

The HTML page is responsive in the sense that when the page width is smaller than 760px, sidenotes and margin notes will be hidden by default. For sidenotes, you can click their numbers (the superscripts) to toggle their visibility. For margin notes, you may click the circled plus signs to toggle visibility.

# More Examples

The rest of this document consists of a few test cases to make sure everything still works well in slightly more complicated scenarios. First we generate two plots in one figure environment with the chunk option `fig.show = 'hold'`:

```{r fig-two-together, fig.cap="Two plots in one figure environment.", fig.show='hold', cache=TRUE, message=FALSE}
p <- ggplot(mtcars2, aes(hp, mpg, color = am)) +
  geom_point()
p
p + geom_smooth()
```

Then two plots in separate figure environments (the code is identical to the previous code chunk, but the chunk option is the default `fig.show = 'asis'` now):

```{r fig-two-separate, ref.label='fig-two-together', fig.cap=sprintf("Two plots in separate figure environments (the %s plot).", c("first", "second")), cache=TRUE, message=FALSE}
```

You may have noticed that the two figures have different captions, and that is because we used a character vector of length 2 for the chunk option `fig.cap` (something like `fig.cap = c('first plot', 'second plot')`).

Next we show multiple plots in margin figures. Similarly, two plots in the same figure environment in the margin:

```{r fig-margin-together, fig.margin=TRUE, fig.show='hold', fig.cap="Two plots in one figure environment in the margin.", fig.width=3.5, fig.height=2.5, cache=TRUE}
p
p + geom_smooth(method = 'lm')
```

Then two plots from the same code chunk placed in different figure environments:

```{r fig-margin-separate, fig.margin=TRUE, fig.cap=sprintf("Two plots in separate figure environments in the margin (the %s plot).", c("first", "second")), fig.width=3.5, fig.height=2.5, cache=TRUE}
knitr::kable(head(iris, 15))
p
knitr::kable(head(iris, 12))
p + geom_smooth(method = 'lm')
knitr::kable(head(iris, 5))
```

We blended some tables in the above code chunk only as *placeholders* to make sure there is enough vertical space among the margin figures, otherwise they will be stacked tightly together. For a practical document, you should not insert too many margin figures consecutively and make the margin crowded.

You do not have to assign captions to figures. We show three figures with no captions below in the margin, in the main column, and in full width, respectively.

```{r fig-nocap-margin, fig.margin=TRUE, fig.width=3.5, fig.height=2, cache=TRUE}
# a boxplot of weight vs transmission; this figure
# will be placed in the margin
ggplot(mtcars2, aes(am, wt)) + geom_boxplot() +
  coord_flip()
```

```{r fig-nocap-main, cache=TRUE}
# a figure in the main column
p <- ggplot(mtcars, aes(wt, hp)) + geom_point()
p
```

```{r fig-nocap-fullwidth, fig.fullwidth=TRUE, fig.width=10, fig.height=3, cache=TRUE}
# a fullwidth figure
p + geom_smooth(method = 'lm') + facet_grid(~ gear)
```

# Some Notes on Tufte CSS

There are a few other things in Tufte CSS that we have not mentioned so far. If you prefer `r sans_serif('sans-serif fonts')`, use the function `sans_serif()` in **tufte**. For epigraphs, you may use a pair of underscores to make the paragraph italic in a block quote, e.g.

> *I can win an argument on any topic, against any opponent. People know this, and steer clear of me at parties. Often, as a sign of their great respect, they don't even invite me.*
>
> `r quote_footer('--- Dave Barry')` We hope you will enjoy the simplicity of R Markdown and this R package, and we sincerely thank the authors of the Tufte-CSS and Tufte-LaTeX projects for developing the beautiful CSS and LaTeX classes. Our **tufte** package would not have been possible without their heavy lifting.

You can turn on/off some features of the Tufte style in HTML output. The default features enabled are:

``` yaml
output:
  tufte::tufte_html:
    tufte_features: ["fonts", "background", "italics"]
```

If you do not want the page background to be lightyellow, you can remove `background` from `tufte_features`. You can also customize the style of the HTML page via a CSS file. For example, if you do not want the subtitle to be italic, you can define

``` css
h3.subtitle em {
  font-style: normal;
}
```

in, say, a CSS file `my_style.css` (under the same directory of your Rmd document), and apply it to your HTML output via the `css` option, e.g.,

``` yaml
output:
  tufte::tufte_html:
    tufte_features: ["fonts", "background"]
    css: "my_style.css"
```

There is also a variant of the Tufte style in HTML/CSS named "[Envisoned CSS](https://github.com/nogginfuel/envisioned-css)". This style can be used by specifying the argument `tufte_variant = 'envisioned'` in `tufte_html()`[^5], e.g.

[^5]: The actual Envisioned CSS was not used in the **tufte** package. We only changed the fonts, background color, and text color based on the default Tufte style.

``` yaml
output:
  tufte::tufte_html:
    tufte_variant: "envisioned"
```

To see the R Markdown source of this example document, you may follow [this link to Github](https://github.com/rstudio/tufte/raw/main/inst/rmarkdown/templates/tufte_html/skeleton/skeleton.Rmd), use the wizard in RStudio IDE (`File -> New File -> R Markdown -> From Template`), or open the Rmd file in the package:

```{r eval=FALSE}
file.edit(
  tufte:::template_resources(
    'tufte_html', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

This document is also available in [Chinese](https://rstudio.github.io/tufte/cn/), and its `envisioned` style can be found [here](https://rstudio.github.io/tufte/envisioned/).

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
